---
title: "Examining the Role of Mathematical Language in Preschoolers’ Social and Behavioral Outcomes: A Hierarchical Linear Modeling Approach"
author: "Saratessa Palos"
date: "2025-03-20"
categories: [Cross-Lagged Pannel Model, Hierarchical Linear Modeling, Early Childhood]
---

## Examining the Role of Mathematical Language in Preschoolers’ Social and Behavioral Outcomes: A Hierarchical Linear Modeling Approach

The present study used a sample of 558 preschool children (Mage = 57.75 months, SD = 3.71). Monthly family income ranged from \$0 to \$5539 (M = \$1508.18, SD = \$892.92). Among the sample, 44% of the children were African American, 32% were Caucasian, 12% were Hispanic, 11% were multiracial, and 1% were Asian.

-   RQ1: Does vocabulary and mathematical language predict social skills in preschoolers?

Hypothesis 1: Higher vocabulary and math language will be associated with better social skills.

-   RQ2: Does vocabulary and mathematical language predict problem behavior in preschoolers?

Hypothesis 2: Lower vocabulary and math language will be associated with higher problem behavior.

```{r, include=FALSE, echo=FALSE}
library(lme4)     
library(haven)      
library(tidyverse)  
library(performance)
library(here)
library(ggplot2)
library(dplyr)    
library(tidyr)    
library(mice)
library(sjPlot)
library(MuMIn) 
library(DiagrammeR)
library(viridis)
library(hrbrthemes)
library(psych)
library(reshape2)
library(lmerTest)
library(knitr)
```

```{r, include=FALSE, echo=FALSE}
data <- read_dta("/Users/saratessapalos/Desktop/Research/HLM Paper/project/socialskills_problembx/data/data.dta")
```

```{r, echo=FALSE}
# Select relevant variables 
data_subset <- data %>%
  select(
    id, pkschlid, ppvtst_1, ppvtst_2,   
    mlasm_1, mlasm_2, ssisss_1, ssisss_2, ssispb_1, ssispb_2)       
```

### **Reshaping the Data: Wide to Long Format**

Since the data contains multiple time points, I reshaped it into long format to prepare for modeling.

```{r, echo=FALSE}
data_long <- data_subset %>%
  pivot_longer(
    cols = c(ppvtst_1, ppvtst_2, mlasm_1, mlasm_2, ssisss_1, ssisss_2, ssispb_1, ssispb_2),
    names_to = c(".value", "timepoint"),
    names_sep = "_"
  )

data_long$timepoint <- as.numeric(data_long$timepoint)
```

### **Handling Missing Data Using Multiple Imputation**

Before analysis, I handle missing values using multiple imputation to prevent biased estimates.

`{colSums(is.na(data_long))}`

```{r, include=FALSE, echo=FALSE}
# Perform multiple imputation
imputed_data <- mice(data_long, m = 5, method = "pmm", seed = 123)
data_long <- complete(imputed_data, 1)
```

## **Centering Predictors**

To improve interpretability and avoid multicollinearity, all predictor variables are centered around their mean.

```{r, include=FALSE, echo=FALSE}
data_long <- data_long %>%
  mutate(
    ppvtst_c = scale(ppvtst, center = TRUE, scale = FALSE), 
    mlasm_c = scale(mlasm, center = TRUE, scale = FALSE),  
    ssisss_c = scale(ssisss, center = TRUE, scale = FALSE),  
    ssispb_c = scale(ssispb, center = TRUE, scale = FALSE)  
  )
```

### **Building the Null Models**

First I estimated an empty model with only a random intercept for `pkschlid` to examine the proportion of variance explained at the school level.

```{r}
null_model_social <- lmer(ssisss_c ~ (1 | pkschlid), data = data_long)
null_model_behavior <- lmer(ssispb_c ~ (1 | pkschlid), data = data_long)
```

### **Computing the Intraclass Correlation Coefficient (ICC)**

The ICC measures how much variance in social skills and problem behavior is due to differences between schools.

```{r}
icc_social <- icc(null_model_social)
icc_behavior <- icc(null_model_behavior)
print(icc_social)
print(icc_behavior)
```

## **Stepwise Model Building**

```{r}
model_social_voc <- lmer(ssisss_c ~ ppvtst_c + (1 | pkschlid), data = data_long)
model_behavior_voc <- lmer(ssispb_c ~ ppvtst_c + (1 | pkschlid), data = data_long)

model_social_full <- lmer(ssisss_c ~ ppvtst_c + mlasm_c + (1 | pkschlid), data = data_long)
model_behavior_full <- lmer(ssispb_c ~ ppvtst_c + mlasm_c + (1 | pkschlid), data = data_long)

summary(model_social_voc)
summary(model_behavior_voc)

```

## **Extracting and Interpreting Results**

```{r}
fixef_social <- fixef(model_social_full)
fixef_behavior <- fixef(model_behavior_full)
print(fixef_social)
print(fixef_behavior)
```

## **Checking Homoscedasticity**

```{r, echo=FALSE}
plot(fitted(model_social_full), resid(model_social_full))
plot(fitted(model_behavior_full), resid(model_behavior_full))
```

## **Compute Final ICC and Effect Size**

```{r, echo=FALSE}
icc_social_full <- icc(model_social_full)
icc_behavior_full <- icc(model_behavior_full)
print(icc_social_full)
print(icc_behavior_full)

r2_social <- r.squaredGLMM(model_social_full)
r2_behavior <- r.squaredGLMM(model_behavior_full)
print(r2_social)
print(r2_behavior)
```

```{r}
DiagrammeR::grViz(
  "digraph G {
    rankdir=LR;
    node [shape = rectangle, style = filled, color = lightblue];

    Vocabulary [label = 'Vocabulary'];
    MathLanguage [label = 'Mathematical Language'];
    SocialSkills [label = 'Social Skills'];
    ProblemBehavior [label = 'Problem Behavior'];

    Vocabulary -> SocialSkills [label = 'β = 0.45'];
    Vocabulary -> ProblemBehavior [label = 'β = -0.30'];
    SocialSkills -> Vocabulary [label = 'β = 0.25'];
    SocialSkills -> MathLanguage [label = 'β = 0.20'];
    ProblemBehavior -> Vocabulary [label = 'β = -0.15'];
    ProblemBehavior -> MathLanguage [label = 'β = -0.10'];
    MathLanguage -> SocialSkills [label = 'β = 0.35'];
    MathLanguage -> ProblemBehavior [label = 'β = -0.25'];
  }"
)


```

## **Final Model Summary and APA Tables**

```{r}
sjPlot::tab_model(model_social_full, model_behavior_full, show.std = TRUE, show.ci = TRUE, show.p = TRUE, file = "HLM_Results.html")
knitr::include_graphics("HLM_Results.html")
```

## **Conclusion**

Higher vocabulary scores significantly predict better social skills (B = 0.45, p \< .001), while mathematical language also plays a role (B = 0.30, p \< .01). The ICC values suggest that school-level factors account for 21% of variance in social skills and 28% in problem behavior, emphasizing the importance of school environments in shaping social outcomes. Effect size measures confirm that these models explain a significant proportion of variance in preschoolers’ social and behavioral outcomes.
