---
title: "Exploring Precision Teaching (PT) Math pinpoints"
author: "Saratessa"
date: "2025-02-27"
categories: [Precision Teaching, Math Instruction, Clustering]
---

This following post utilized data from a private precision teaching clinic in the United States. Each student in this data has a score for accuracy and fluency of math pinpoints at intake and then their score for those skills on a CBM after 40 hours of instruction.

This report includes: **Sample visualizations** to understand participant distribution. **Data reshaping** to convert the dataset into a tidy format. **Analysis for three research questions** using clustering, regression, and t-tests.

**###Research Questions and Methods:###**

1.  **What math skills, when grouped, predict other skills?**

    -   **Hierarchical clustering** is used to group skills that show similar learning patterns, helping identify which skills are related to others.

2.  **What math skills predict outcomes on the CBM after 40 hours of instruction?**

    -   **A multiple regression model** is used.

3.  **How do targeted versus untargeted math skills relate to CBM outcomes?**

    -   **ANCOVA** is used.

Before reading in the data, I cleaned it in a different script so that I could remove the names and replace with unique id numbers (to protect student privacy).

In this project, the first step I took was to load all required R packages. For data manipulation (`tidyverse`), visualizations (`ggplot2`), XXXXXXX, and running regression models (`lme4`).

```{r}
library(tidyverse)
library(ggplot2)
library(cluster)
library(broom)
library(lme4)
```

Next I read in the data from my local and then used (`glimpse`) to see an overview of the data.

```{r}
file_path <- "/Users/saratessapalos/Desktop/Research/FIT learning/fitdata/data/Cleaned_Data1.csv"
df <- read_csv(file_path)
```
create visualizations of the sample

```{r}

### Students With vs. Without Diagnosis
df %>% 
  mutate(Has_Diagnosis = ifelse(Diagnosis == "None" | is.na(Diagnosis), "No Diagnosis", "Has Diagnosis")) %>%
  count(Has_Diagnosis) %>%
  ggplot(aes(x = Has_Diagnosis, y = n, fill = Has_Diagnosis)) +
  geom_bar(stat = "identity", color = "black") +
  labs(title = "Students With vs. Without Diagnosis",
       x = "Diagnosis Status",
       y = "Count") +
  scale_fill_manual(values = c("#FFC0CB", "#800080"))

### Grade Level Distribution
df %>%
  filter(!is.na(`Grade Level`)) %>%
  count(`Grade Level`) %>%
  ggplot(aes(x = reorder(`Grade Level`, n), y = n)) +
  geom_bar(stat = "identity", fill = "#800080", color = "black") +
  labs(title = "Grade Level Distribution",
       x = "Grade Level",
       y = "Count") +
  coord_flip()

### Diagnosis Type Count
df %>%
  mutate(Diagnosis_Category = case_when(
    Diagnosis == "Mood Disorder - Anxiety (e.g. selective mutism, anxiety, GAD)" ~ "Mood Disorder",
    Diagnosis == "Learning Disability - Speech and language - Speech delay" ~ "Learning Disability",
    Diagnosis == "None" | is.na(Diagnosis) ~ "None"
  )) %>%
  count(Diagnosis_Category) %>%
  ggplot(aes(x = Diagnosis_Category, y = n, fill = Diagnosis_Category)) +
  geom_col(color = "black") +
  labs(title = "Diagnosis Type Count",
       x = "Diagnosis Type",
       y = "Count") +
  scale_fill_manual(values = c("#FFC0CB", "#800080", "gray"))
```

## Reshape Data to Tidy Format

[edit this now that changes were made\]I reshaped the dataset to a "tidy" format, where each row represents: Participant_ID, Grade Level, Diagnosis, Skill, Measure\` (Intake Accuracy, Post Fluency, etc.), Score. I converted categorical variables to factors to avoid regression errors. And I removed levels with only one category to prevent contrast errors in regression.

```{r}
# Start with clean data (row 4 onwards)
clean_df <- df[4:nrow(df), ]

# Step 2: Create proper column names by combining rows 1 and 2
col_names_skill <- as.character(unlist(df[1, ]))
col_names_assessment <- as.character(unlist(df[2, ]))

# Fix any NA or empty values in column names
col_names_skill <- ifelse(is.na(col_names_skill) | col_names_skill == "", paste0("Column_", 1:length(col_names_skill)), col_names_skill)
col_names_assessment <- ifelse(is.na(col_names_assessment) | col_names_assessment == "", "Unknown", col_names_assessment)

# Create combined names for all columns
full_col_names <- vector("character", length(col_names_skill))
for(i in 1:length(col_names_skill)) {
  if(i <= 3) {
    # Keep the original names for the first three columns
    full_col_names[i] <- col_names_skill[i]
  } else {
    # Combine skill and assessment type for data columns
    full_col_names[i] <- paste(col_names_skill[i], col_names_assessment[i], sep = "_")
  }
}

# Make sure all column names are unique
if(any(duplicated(full_col_names))) {
  # Add a suffix to duplicated names
  dups <- which(duplicated(full_col_names))
  for(i in dups) {
    full_col_names[i] <- paste0(full_col_names[i], "_", i)
  }
}

# Assign the new column names
colnames(clean_df) <- full_col_names

# Step 3: Convert to tidy format
tidy_df <- clean_df %>%
  # Convert character columns to numeric (except first 3 columns)
  mutate(across(4:ncol(.), ~as.numeric(as.character(.)))) %>%
  # Convert to long format
  pivot_longer(
    cols = 4:ncol(.),
    names_to = "Skill_Measure",
    values_to = "Score"
  ) %>%
  # Split Skill_Measure into Skill and Measure
  # Use a more flexible pattern to handle different formats
  separate(Skill_Measure, 
           into = c("Skill", "Measure"),
           sep = "_(?=(Intake|Post|Targeted)($|_))",  # Split before Intake, Post, or Targeted
           fill = "right") %>%  # Handle any missing values
  # Remove any trailing/leading whitespace
  mutate(
    Skill = trimws(Skill),
    Measure = trimws(Measure)
  )
```

```{r}
# Check the number of NAs in the Score column
na_count <- sum(is.na(tidy_df$Score))
total_count <- nrow(tidy_df)
na_percentage <- (na_count / total_count) * 100

# Print summary
cat("Total rows:", total_count, "\n")
cat("NA values in Score:", na_count, "\n")
cat("Percentage NA:", na_percentage, "%\n")

# See what the tidy data looks like
head(tidy_df)
```

////// \## Answer Research Questions

### **Q1: What math skills, when grouped, predict other skills?**

to address this I used a **Correlation analysis** to identify skills that **improve together**. Then I did **Association rule mining** to find patterns where mastering one skill predicts mastery of another. I then created the following visualizations: a **correlation heatmap** and **association rule graph**.

```{r}
# Create a correlation matrix for skill scores


# Visualize correlation matrix

```
